Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach

### 2014-05-23 manuscripts/manuscript.tex
updates as sent to steve

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Avoiding tipping points in ecosystem based fisheries management: a non-parametric Bayesian approach},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{Avoiding tipping points in ecosystem based fisheries management: a
non-parametric Bayesian approach}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig@gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[cstar]{Marc Mangel}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric
Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}

  \begin{abstract}
  Model uncertainty and limited data coverage are fundamental challenges
  to robust ecosystem management. These challenges are acutely highlighted
  by concerns that many ecological systems may contain tipping points.
  Before a collapse, we do not know where the tipping points lie, if they
  exist at all. Hence, we know neither a complete model of the system
  dynamics nor do we have access to data in some large region of
  state-space where such a tipping point might exist. These two sources of
  uncertainty frustrate state-of-the-art parametric approaches to decision
  theory and optimal control. Here we illustrate how a non-parametric
  approach using a Gaussian process prior provides a more flexible
  representation of this inherent uncertainty. We can then embed the
  Gaussian process in a stochastic dynamic programming framework in order
  to make robust management predictions under both model uncertainty and
  limited data. We conduct a set of simulations to evaluate this approach
  as compared with the standard approach of using model selection to
  choose from a set of candidate models. In keeping with previous results,
  we find that model selection erroneously favors models without tipping
  points -- leading to harvest policies that guarantee extinction. The
  Gaussian process approach performs nearly as well as the true model and
  significantly outperforms standard appraoches. We illustrate this using
  examples of simulated single-species dynamics, where the standard model
  selection approach should be most effective, and find that it still
  fails to account for uncertainty appropriately and leads to population
  crashes, while management based on the Gaussian process approach does
  not. We discuss opportunities to adapt this approach to multiple species
  models in further work.
  \end{abstract}
   \begin{keyword} Bayesian \sep Structural Uncertainty \sep Nonparametric \sep Optimal Control \sep Decision Theory \sep Gaussian Processes \sep Fisheries Management \sep \end{keyword}
 \end{frontmatter}

@@

Decision making under uncertainty is a ubiquitous challenge of natural
resource management and conservation. Ecological dynamics are frequently
complex and difficult to measure, making uncertainty in our
understanding a persistent challenge to effective management.
Decision-theoretic approaches provide a framework to determine the best
sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified (Fischer \emph{et al.} 2009).
Concerns about the potential sudden collapse of fisheries and other
ecosystems (Worm \emph{et al.} 2006; Scheffer \emph{et al.} 2009) has
increasingly emphasized the difficulties of formulating even
qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other conservation and resource management problems. Economic value
and ecological concern have made marine fisheries the crucible for much
of the founding work (Gordon 1954; May \emph{et al.} 1979; Reed 1979;
Ludwig \& Walters 1982) on management under uncertainty. Global declines
(Worm \emph{et al.} 2006) and the controversy surrounding their
interpretation (Hilborn 2007; Worm \emph{et al.} 2009) have made
understanding these challenges all the more pressing.

Uncertainty enters the decision-making process at many levels: intrinsic
stochasticity in biological processes, measurements, and implementation
of policy (\emph{e.g.} Reed 1979; Clark \& Kirkwood 1986; Roughgarden \&
Smith 1996; Sethi \emph{et al.} 2005), parametric uncertainty
(\emph{e.g.} Ludwig \& Walters 1982; Hilborn \& Mangel 1997; McAllister
1998; Schapaugh \& Tyre 2013), and model or structural uncertainty
(\emph{e.g.} Williams 2001; Cressie \emph{et al.} 2009; Athanassoglou \&
Xepapadeas 2012). Of these, structural uncertainty is generally the
hardest to quantify. Typical approaches assume a weak notion of model
uncertainty in which the correct model (or reasonable approximation) of
the dynamics must be identified from among a handful of alternative
models. Here we consider an approach that addresses uncertainty at each
of these levels without assuming the dynamics follow a particular
(i.e.~parametric) structure.

Often we do not understand the system dynamics well enough to know if a
model provides a good approximation over the relevant range of states
and timescales (criteria that we loosely refer to as defining the
``right'' or ``true'' model.)\\Typical solutions involve either model
choice, model averaging, or introducing yet greater model complexity of
which others may be special cases (model averaging being one such way to
construct such a model) (Williams 2001; Cressie \emph{et al.} 2009;
Athanassoglou \& Xepapadeas 2012). Even setting aside other
computational and statistical concerns (e.g. (Cressie \emph{et al.}
2009)), these approaches do not address our second concern -
representing uncertainty outside the observed data range.

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated.
This extrapolation uncertainty is felt most keenly in decision-theoretic
(or optimal control) applications, since (a) exploring the potential
action space typically involves considering actions that may move the
system outside the range of observed behavior, and (b)
decision-theoretic algorithms rely not only on reasonable estimates of
the expected outcomes, but depend on the weights given to all possible
outcomes (\emph{e.g.} Weitzman 2013).

This difficult position of having neither the true model nor data that
covers the full range of possible states is unfortunately the rule more
than the exception. The potential concern of tipping points in
ecological dynamics (Scheffer \emph{et al.} 2001; Polasky \emph{et al.}
2011) reflects these concerns -- as either knowledge of the true model
or more complete sampling of the state space would make it easy to
identify if a tipping point existed. If we do not know but cannot rule
out such a possibility, then we face decision-making under this dual
challege of model uncertainty and incomplete data coverage.

These dual concerns pose a substantial challenge to existing
decision-theoretic approaches (Brozović \& Schlenker 2011). Because
intervention is often too late after a tipping point has been crossed
(but see Hughes et al. (2013)), management is most often concerned with
avoiding potentially catastrophic tipping points before any data is
available at or following a transition that would more clearly reveal
these regime shift dynamics (e.g. Bestelmeyer \emph{et al.} 2012).

Here we illustrate how a stochastic dynamic programming (SDP) algorithm
(Mangel \& Clark 1988; Marescot \emph{et al.} 2013) can be driven by the
predictions from a Bayesian non-parametric (BNP) approach (Munch
\emph{et al.} 2005a). This provides two distinct advantages compared
with contemporary approaches. First, using a BNP sidesteps the need for
an accurate model-based description of the system dynamics. Second, the
BNP can better reflect uncertainty that arises when extrapolating a
model outside of the data on which it was fit. We illustrate that when
the correct model is not known, this latter feature is crucial to
providing a robust decision-theoretic approach in face of substantial
structural uncertainty.

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics. In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a more
state-space dependent representation of uncertainty. This permits a much
greater uncertainty far from the observed data than near the observed
data. These features allow the GPDP approach to find robust management
solutions in face of limited data and without knowledge of the correct
model structure.

The idea that any approach can perform well without either having to
know the model or have particularly good data should immediately draw
suspicion. The reader must bear in mind that the strength of our
approach comes not from black-box predictive power from such limited
information, but rather, by providing a more honest expression of
uncertainty outside the observed data without sacrificing the predictive
capacity near the observed data. By coupling this more accurate
description of what is known and unknown to the decision-making under
uncertainty framework provided by stochastic dynamic programming, we are
able to obtain more robust management policies than with common
parametric modeling approaches.

The nature of decision-making problems provides a compelling and
pragmatic way to compare models: Rather than compare models in terms of
best fit to data, model performance is defined in the concrete terms of
the decision-maker's objectives. Although much argument can be made over
the `correct' objective function, e.g.~how to account for the social
value of fish left in the sea vs.~the commercial value of fish
harvested; see Halpern et al. (2013) for further discussion of this
issue. We can always compare model performance across multiple potential
objectives. The important point is that a decision-maker does not
necessarily need a model that provides the best mechanistic
understanding or the best long-term outcome, but rather the one that
idelivers the best management performance over the period of interest.

\subsection{Background on the Gaussian
Process}\label{background-on-the-gaussian-process}

We use a Bayesian nonparameteric regression approach based on Gaussian
process priors. The use of Gaussian process (GP) regression (or
``kriging'' in the geospatial literature) to formulate a predictive
model is relatively new in the context of modeling dynamical systems
(Kocijan \emph{et al.} 2005), and was first introduced in the context
ecological modeling and fisheries management in Munch et al. (2005b). GP
models have subsequently been used to test for the presence of Allee
effects (Sugeno \& Munch 2013a), estimate the maximum reproductive rate
(Sugeno \& Munch 2013b), determine temporal variation in food
availability (Sigourney \emph{et al.} 2012), and provide a basis for
identifying model-misspecification (Thorson \emph{et al.} 2014). An
accessible and thorough introduction to the formulation and use of GPs
can be found in Rasmussen \& Williams (2006). The nonparametric moniker
for the GP approach is unfortunate as (a) it still involves the
estimation of parameters, and (b), statisticians also use non-parametric
to mean models that make only weak distributional assumptions (e.g.
Lehmann 1975). However, we retain the terminology for consistency with
previous studies.

@@

\subsubsection{Statement of the optimal control
problem}\label{statement-of-the-optimal-control-problem}

To illustrate the application of the Gaussian process-based dynamic
programming approach (hereafter GPDP) we focus on the classical problem
of determining a harvest policy that will maximize the expected
discounted profit over a fixed time interval (Reed 1979; Ludwig \&
Walters 1982; Mangel \& Clark 1988). For simplicity we assume profit is
proportional to the realized harvest.\\For comparison, we consider the
performance of the GPDP policy against policies derived under several
alternative parametric models.

We focus on a class of one-dimensional models of population dynamics as
a `best-case' scenario for the parametric model performance relative to
the Gaussian process. We consider this a best-case scenario for the
following reasons: first, the underlying dynamics are always simuated
from one the parametric models in the set, whereas in reality we never
have the ``true'' model among our candidates. By choosing
one-dimensional models with few parameters, we limit the chance that
poor performance will be due to our inability to estimate parameters
well, something that becomes a more severe problem for
higher-dimensional parametric models. Further, the parametric models we
consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable
states.

@@

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model (Allen \& Tanner 2005), the
Myers model (Myers \emph{et al.} 1995), equations (1)-(3), which we
parameterize as follows:

We consider three parametric models that relate population size at time
$t+1$, $X_{t+1}$, to population size at time $t$, $X_t$ and harvest at
time $t$, $h_t$ by

\[ X_{t+1}= Z_t f(S_t,p) \]

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $p$ is a vector of parameters, and $S_t$ the escapement (Reed 1979)

\[S_t = X_t - h_t\]

or the Ricker model,

@@

where $K$ is the carrying capacity and $r$ the growth rate. For the
Allen model,

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \[ f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - X_C\right)} \]
\end{enumerate}

Where $r$ and $K$ are as before, and $X_c$ denotes the location of the
unstable (tipping) point. For the Myers model,

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{2}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \[ f(S(t) | p)  = \frac{r S_t^{\theta}}{1 - \frac{S_t^\theta}{K}}\]
\end{enumerate}

Where $r$ and $K$ as before, and $\theta$ the nonlinearity of the Myers
model (corresponding to Beverton-Holt dynamics for $\theta$ = 1 and an
Allee effect with alternative stable states exist at $\theta > 2$).

The familiar Ricker model involves two parameters, corresponding to a
growth rate and a carrying capacity, and cannot support alternative
stable growth rate dynamics. The Allen model resembles the Ricker
dynamics with an added Allee effect parameter (Courchamp \emph{et al.}
2008), below which the population cannot persist. The Myers model also
has three parameters and contains an Allee threshold, but has
compensatory rather than over-compensatory density dependence
(resembling a Beverton-Holt curve rather than a Ricker curve at high
densities.)\\We assume multiplicative log-normal noise perturbs the
growth predicted by the each of the deterministic model skeletons
described above. This introduces one additional parameter $\sigma$ that
must be estimated by each model.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model. The
Allen model is thus structurally correct and is expected to provide a
best-case scenario when the `true' dynamics are known. The Ricker model
is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

We consider a period of 40 years of training data: long enough that the
estimates are not dependent on the particular realization, while longer
times are not likely to provide substantial improvement (i.e.~the
results are not sensitive to this interval). Each of the models
(described below) is fit to the same training data, as shown in Figure
1.

We inferred posterior distributions for the parameters of each model in
a Bayesian context using Gibbs sampling (Gelman et al. (2003)
implemented in R (R Core Team 2013) using jags, (Su \& Masanao Yajima
2013)). We choose uninformative uniform priors for all parameters (See
appendix Tables S1-S3; R code provided). We show one-step-ahead
predictions of these model fits in Figure 1. Each chain was tested for
Gelman-Rubin convergence and results were robust to longer runs. For
each simulation we also applied several commonly used model selection
criteria (AIC, BIC, DIC, see Burnham \& Anderson (2002)) to identify the
best fitting model.

An optimal policy function was then inferred through stochastic dynamic
programming for each model given the posterior distributions of the
parameter estimates. This policy maximizes the expectation of the value
function integrated over the parameter uncertainty. (code implementing
this algorithm provided in the Appendix).

@@

@@

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a SDP algorithm. Recall that unlike
the parametric models the Gaussian process with fixed parameters already
predicts a distribution of curves rather than a single curve. We must
first integrate over this distribution of curves given a sampling of
parameter values drawn from the posterior distribution of the two GP
parameters, before integrating over the posterior of those parameters
themselves.

@@

@@

\subsection{Gaussian process model}\label{gaussian-process-model}

Assuming the data are known with some process noise,

$\varepsilon$ IID normal, variance $\sigma_n^2$. Then under the GP,

>>>>

\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \usepackage{fontspec}
  \ifxetex
    \usepackage{xltxtra,xunicode}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use microtype if available
\IfFileExists{microtype.sty}{\usepackage{microtype}}{}
\usepackage{longtable}
\usepackage{graphicx}
% We will generate all images so they have a width \maxwidth. This means
% that they will get their normal width if they fit onto the page, but
% are scaled down if they would overflow the margins.
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth
\else\Gin@nat@width\fi}
\makeatother
\let\Oldincludegraphics\includegraphics
\renewcommand{\includegraphics}[1]{\Oldincludegraphics[width=\maxwidth]{#1}}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={},
            pdftitle={Avoiding tipping points in fisheries management: A Bayesian Non-Parametric approach},
            colorlinks=true,
            urlcolor=blue,
            linkcolor=magenta,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}
% Pandoc toggle for numbering sections (defaults to be off)
\setcounter{secnumdepth}{0}
% Pandoc header

@@

  \title{Avoiding tipping points in fisheries management: A Bayesian
Non-Parametric approach}
    \author[cstar]{Carl Boettiger\corref{c1}}
   \ead{cboettig@gmail.com} 
   \cortext[c1]{Corresponding author}
    \author[cstar]{Marc Mangel}

@@

      \address[cstar]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA}    
    \address[marc]{Center for Stock Assessment Research, Department of Applied Math and
Statistics, University of California, Mail Stop SOE-2, Santa Cruz, CA
95064, USA and Department of Biology, University of Bergen, Bergen,
Norway 9020}    
    \address[noaa]{Southwest Fisheries Science Center, National Oceanic and Atmospheric
Administration, 110 Shaffer Road, Santa Cruz, CA 95060, USA}

  \begin{abstract}
  Model uncertainty and limited data coverage are fundamental challenges
  to robust management of human intervention in a natural system. These
  challenges are acutely highlighted by concerns that many ecological
  systems may contain tipping points, such as Allee population
  sizes.\\Before a collapse, we do not know where the tipping points lie,
  if they exist at all. Hence, we know neither a complete model of the
  system dynamics nor do we have access to data in some large region of
  state-space where such a tipping point might exist. We illustrate how a
  Bayesian Non-Parametric (BNP) approach using a Gaussian Process (GP)
  prior provides a flexible representation of this inherent uncertainty.
  We embed GPs in a stochastic dynamic programming framework in order to
  make robust management predictions with both model uncertainty and
  limited data. We conduct a set of simulations to evaluate this approach
  as compared with the standard approach of using model selection to
  choose from a set of candidate models. We find that model selection
  erroneously favors models without tipping points -- leading to harvest
  policies that guarantee extinction. The GP approach performs nearly as
  well as the true model and significantly outperforms standard
  appraoches. We illustrate this using examples of simulated
  single-species dynamics, where the standard model selection approach
  should be most effective, and find that it still fails to account for
  uncertainty appropriately and leads to population crashes, while
  management based on the GP approach does not, since it does not
  underestimate the uncertainty outside of the observed data.
  \end{abstract}
   \begin{keyword} Bayesian \sep Structural Uncertainty \sep Nonparametric \sep Optimal Control \sep Decision Theory \sep Gaussian Processes \sep Fisheries Management \sep \end{keyword}
 \end{frontmatter}

@@

Decision making under uncertainty is a ubiquitous challenge in the
management of human intervention in natural resources and conservation.
Ecological dynamics are frequently complex and difficult to measure,
making uncertainty in our understanding a persistent challenge.
Decision-theoretic approaches provide a framework to determine the best
sequence of actions in face of uncertainty, but only when that
uncertainty can be meaningfully quantified (Fischer et al. 2009). Over
the last four decades (but stretching even further back in time), marked
by the seminal contributions of Clark (1976); Clark (2009) and Walters
and Hilborn (1978), dynamic optimization methods, particularly
Stochastic Dynamic Programming (SDP, Clark (2009)), have become
increasingly important as a means of understanding how to management
human intervention into natural systems. Concomittantly, there has been
increasing recognition of the importance of Allee population sizes or
`tipping points' (Scheffer et al. 2001, Polasky et al. 2011) that
management actions may inadvertently cause populations to cross.
Scheffer et al. (2009) emphasized the difficulties of formulating even
qualitatively correct models of the underlying processes.

We develop a novel approach to address these concerns in the context of
fisheries; though the underlying challenges and methods are germane to
many other problems of conservation or natural resource exploitation,
such as forestry. Economic value and ecological concern have made marine
fisheries the crucible for much of the founding work (Gordon 1954, Clark
1976, 2009, May et al. 1979, Reed 1979, Ludwig and Walters 1982) on
management under uncertainty. Global declines (Worm et al. 2006) and the
controversy surrounding their interpretation (Hilborn 2007, Worm et al.
2009) have made understanding these challenges all the more pressing.

Uncertainty enters the decision-making process at many levels: intrinsic
stochasticity in biological processes, measurements, and implementation
of policy (\emph{e.g.} Reed 1979, Clark and Kirkwood 1986, Roughgarden
and Smith 1996, Sethi et al. 2005), parametric uncertainty (\emph{e.g.}
Ludwig and Walters 1982, Hilborn and Mangel 1997, McAllister 1998,
Schapaugh and Tyre 2013), and model or structural uncertainty
(\emph{e.g.} Williams 2001, Cressie et al. 2009, Athanassoglou and
Xepapadeas 2012). Of these, structural uncertainty is generally the
hardest to quantify. Typical approaches assume a weak notion of model
uncertainty in which the correct model (or reasonable approximation) of
the dynamics must be identified from among a handful of alternative
models, using either model choice, model averaging, or introducing yet
greater model complexity of which others may be special cases (model
averaging being one such way to construct such a model) (Williams 2001,
Cressie et al. 2009, Athanassoglou and Xepapadeas 2012). Even setting
aside other computational and statistical concerns (e.g. Cressie et al.
(2009)), these approaches do not address our second concern -
representing uncertainty outside the observed data range.

Model uncertainty is particularly insidious when model predictions must
be made outside of the range of data on which the model was estimated,
and models may be abused (Mangel et al. 2001). Extrapolation uncertainty
is felt most keenly in decision-theoretic (or optimal control)
applications, since (a) exploring the potential action space typically
involves considering actions that may move the system outside the range
of observed behavior, and (b) decision-theoretic algorithms rely not
only on reasonable estimates of the expected outcomes, but depend on the
weights given to all possible outcomes (\emph{e.g.} Weitzman 2013).

The dual concerns of model uncertainty and incomplete data coverage pose
a substantial challenge to existing decision-theoretic approaches
(Brozović and Schlenker 2011). Because intervention is often (but not
always, see Hughes et al. (2013)) too late after a tipping point has
been crossed, management is most often concerned with avoiding
potentially catastrophic tipping points before any data is available at
or following a transition that would more clearly reveal these regime
shift dynamics (e.g. Bestelmeyer et al. 2012).

We illustrate how a stochastic dynamic programming (SDP) algorithm
(Mangel and Clark 1988, Marescot et al. 2013) can be driven by the
predictions from a Bayesian Non-Parametric (BNP) model of population
dynamics (Munch et al. 2005a). This provides two distinct advantages
compared with contemporary approaches. First, using a BNP sidesteps the
need for an accurate model-based description of the system dynamics.
Second, the BNP can better reflect uncertainty that arises when
extrapolating a model outside of the data on which it was fit. We
illustrate that when the correct model is not known, this latter feature
is crucial to providing a robust decision-theoretic approach in face of
substantial structural uncertainty.

This paper represents the first ecological application of the SDP
decision-making framework without an a priori model of the underlying
dynamics. In contrast to parametric models which can only reflect
uncertainty in parameter estimates, the BNP approach provides a more
state-space dependent representation of uncertainty. This permits a much
greater uncertainty far from the observed data than near the observed
data. These features allow the Gaussian Process Dynamic Programming
(GPDP) approach to find robust management solutions in face of limited
data and without knowledge of the correct model structure.

To illustrate GPDP, we consider the performance of the GPDP policy
against the policies derived under several alternative parametric models
(Reed 1979, Ludwig and Walters 1982, Mangel and Clark 1988). The nature
of decision-making problems provides a compelling and pragmatic way to
compare models: Rather than compare models in terms of best fit to data,
we define model performance in the concrete terms of the
decision-maker's objectives.

@@

We first describe the requirements of dynamic optimization for the
management of human intervention in natural resource systems. After that
we describe three parametric models for population dynamics and the GP
description of population dynamics.

\subsubsection{Requirements of dynamic
optimization}\label{requirements-of-dynamic-optimization}

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. In this
paper, we consider only a single state variable, so that the ideas can
be described as simply as possible. That is, we focus on a class of
one-dimensional models of population dynamics as a `best-case' scenario
for the parametric model performance relative to the GP. This is
abest-case scenario for the parametric models because the underlying
dynamics are always simuated from one the parametric models in the set,
whereas in reality we never have the ``true'' model among our
candidates. By choosing one-dimensional models with few parameters, we
limit the chance that poor performance will be due to our inability to
estimate parameters well, something that becomes a more severe problem
for higher-dimensional parametric models. Further, the parametric models
we consider are those most commonly used in modeling stock-recruitment
dynamics or to model sudden transitions between alternative stable
states.

@@

We let $X_t$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are

\[ X_{t+1}= Z_t f(X_t, h_t, p) \]

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $p$ is a vector of parameters to be estimated from the data. We
describe three choices for $f(X_t,p)$ in the next section.In this simple
model, the control action is a harvest or take, $h_t$, measured in the
same units as $X$, at time $t$. Thus, in the presence of take, the
population size on the right hand side of Eqn 1 is replaced by
$S_t=X_t-h_t$.

To construct the value function, we consider a return when $X_t=x$ and
harvest $h_t=h$ denoted as the reward, $R(x,h)$. For example, if the
return is simply the harvest at time $t$, then $R(x,h)=min(x,h)$. We
assume that future harvests are discounted relative to current ones at a
constant rate of discount $\delta$ and ask for the harvest policy that
maximizes total discounted harvest between the current time $t$ and a
final time $T$, that is we seek to maximize over choices of harvest
$\mathbf{E}_{X_t+1} [ \sum_{t = 0}^{T}  R_t(X_t, h_t) \delta^t]$, where
the state dynamics are given by Eqn 1 and $\mathbf{E}$ denotes the
expectation over the process stochasticity in the future population
state.

In order to find that policy, we introduce the value function$V_t(x_t)$
representing the total discounted catch from time $t$ onwards given that
$X_t=x$. This value function satisfies an equation of Stochastic Dynamic
Programming (SDP, Mangel and Clark 1988, Clark and Mangel 2000, Clark
2009), sometimes called the Bellman-Hamilton-Jacobi equation (Mangel
2014).

\begin{equation}
V_t(x) = \max_{h_t} \left[R(h_t, x_t) + \delta \cdot \mathbf{E}_{X_t+1} V_{t+1}( Z_{t+1} f(x - h_t)) \right]
\end{equation}

That is, at time $t$, when population size is $x$ and harvest $h_t$ is
applied, the immediate return is $R(h_t, x_t)$ and the population size
after the take is $x-h_t$, which is then translated into $X_{t+1}$ by
Eqn 1. We consider the finite time problem with \$T=\$1000, which we
solve using value iteration algorithm (Mangel and Clark 1988, Clark and
Mangel 2000).

We consider three candidate parametric models of the stock-recruitment
dynamics: The Ricker model, the Allen model (Allen and Tanner 2005), the
Myers model (Myers et al. 1995), equations (1)-(3), which we
parameterize as follows for $f(S_t,p)$ in Eqn 1. In all three, we let
$K$ denote the carrying capacity and $r$ the maximum per capita growth
rate. In the Ricker model,

@@

In the Allen model,

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \[ f(S(t)|p) = S_t e^{r \left(1 - \frac{S_t}{K}\right)\left(S_t - X_c\right)} \]
\end{enumerate}

Where $r$ and $K$ are as before, and $X_c$ denotes the location of the
unstable steady state (i.e., the tipping point).

In the Myers model,

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{2}
\itemsep1pt\parskip0pt\parsep0pt
\item
  \[ f(S(t) | p)  = \frac{r S_t^{\theta}}{1 + \frac{S_t^\theta}{K}}\]
\end{enumerate}

Where $\theta = 1$ corresponding to Beverton-Holt dynamics and $\theta$
\textgreater{} 2 leads to Allee effects and multiple stable states.

The Ricker model involves two parameters, corresponding to a growth rate
and a carrying capacity, and cannot support alternative stable growth
rate dynamics. The Allen model resembles the Ricker dynamics with an
added Allee effect parameter (Courchamp et al. 2008), below which the
population cannot persist. The Myers model also has three parameters and
contains an Allee threshold, but has compensatory rather than
over-compensatory density dependence (resembling a Beverton-Holt curve
rather than a Ricker curve at high densities.)

The multiplicative log-normal stochasticity perturbs the growth
predicted by each of these deterministic model skeletons. This
introduces one additional parameter $\sigma$ that must be estimated.

Because of our interest in management performance in the presence of
tipping points, all of our simulations are based on the Allen model. The
Allen model is thus ``structurally correct'' and is expected to provide
a best-case scenario when the `true' dynamics are known. The Ricker
model is a reasonable approximation of these dynamics far from the Allee
threshold (but lacks threshold dynamics), while the Myers model shares
the essential feature of a threshold but differs in the structure.

We consider a period of 40 years of training data: long enough that the
estimates do not depend on the particular realization, while longer
times are not likely to provide substantial improvement (i.e.~the
results are not sensitive to this interval). Each of the models
(described below) is fit to the same training data (Figure 1).

We inferred posterior distributions for the parameters of each model by
Gibbs sampling (Gelman et al. (2003) implemented in R (R Core Team 2013)
using jags, (Su and Masanao Yajima 2013)). We choose uniform priors for
all parameters (See appendix Tables S1-S3; R code provided). We show
one-step-ahead predictions of these model fits in Figure 1. We tested
each chain for Gelman-Rubin convergence and results were robust to
longer runs. For each simulation we also applied several commonly used
model selection criteria (AIC, BIC, DIC, see Burnham and Anderson
(2002)) to identify the best fitting model.

@@

The core difference for our purpose between the GP representation and
the parametric models we have already introduced is the
``non-parametric'' property, which means that an estimated GP model is
defined explicitly in reference to some observed data. Unlike the models
above, it cannot be specified by the value of some parameters alone. The
nonparametric moniker for the GP approach is unfortunate as (a) it still
involves the estimation of parameters, and (b), statisticians also use
non-parametric to mean models that make only weak distributional
assumptions (e.g. Lehmann 1975). However, we retain the terminology for
consistency with previous studies.

The use GP methods to formulate a predictive model is relatively new in
the context of modeling dynamical systems (Kocijan et al. 2005), and was
first introduced in the context ecological modeling and fisheries
management in Munch et al. (2005b). GP models have subsequently been
used to test for the presence of Allee effects (Sugeno and Munch 2013a),
estimate the maximum reproductive rate (Sugeno and Munch 2013b),
determine temporal variation in food availability (Sigourney et al.
2012), and provide a basis for identifying model-misspecification
(Thorson et al. 2014). An accessible and thorough introduction to the
formulation and use of GPs can be found in Rasmussen and Williams
(2006).

A GP is an infinite dimensional (because it considers a continuous
random variable) stochastic process for which any realization is a
multivariate normal distribution. Thus, to characterize the GP we need a
mean function and a covariance matrix. We proceed as follows.

Assuming the data $X_o$ are observed with some process noise,

$\varepsilon$ IID normal, variance $\sigma_n^2$. Then under the GP, the
predicted data $X_p$ given the observations obeys

\[f(X_p|X_o) \sim \mathcal{N}(E,C)\]
\[E = K(X_p, X_o) \left(K(X_o,X_o) - \sigma \mathbb{I} \right)  ^{-1} y\]
\[C = K(X_p, X_p) - K(X_p, X_o) K(X_o,X_o)^{-1} K(X_o, X_p)\]

Where $K$ is the Gaussian process kernel. (Compare to Eqn 2.22 of
Rasmussen and Williams (2006) for a more detailed derivation.)
Throughout, we use the radial basis function kernel:

We use vague inverse Gamma priors on both the lengthscale $\ell$ and
process noise $\sigma$ parameters, of the form

@@

\subsection{The method of Gaussian Process Dynamic
Programming}\label{the-method-of-gaussian-process-dynamic-programming}

It is reasonably straight-forward to derive the harvest policy from the
estimated BP by inserting it into a SDP algorithm. The only difference
is that the uncertainty in the future state under the GP, $X_{t+1}$,
includes both process uncertainty (based on the estimation of $\sigma$)
and ``structural uncertainty'' of the posterior collection of
curves.\\(Recall that given the parameters, a parametric model is
represented by a single curve, while the GP remains a distribution of
curves).\\From the GP posteriors, we can write down the (discritized)
transition matrix representing the probability of going to each state
$X_{t+1}$ given any current state $X_t$ and any harvest $h_t$ (See the
function \texttt{gp\_transition\_matrix()} in the provided R package).
Given this transition matrix, we use the same value iteration algorithm
as in the parametric case to determine the optimal policy.

@@

\subsection{Parametric and GP models for population
dynamics}\label{parametric-and-gp-models-for-population-dynamics}

@@
